---
title: "Foundations for Bayesian Analysis Homework"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

library(tidyverse)
library(stringr)
library(lubridate)
library(kableExtra)
library(cowplot)
library(ggExtra)
library(sfsmisc)
library(janitor)
library(sn)

```

```{r, echo = F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

SalesTrans = read_csv("C:/Users/ellen/Documents/UH/Fall 2020/Class Materials/Section II/Class 1/Data/Sales.csv")
Location = read_csv("C:/Users/ellen/Documents/UH/Fall 2020/Class Materials/Section II/Class 1/Data/Location.csv")
MerGroup = read_csv("C:/Users/ellen/Documents/UH/Fall 2020/Class Materials/Section II/Class 1/Data/MerGroup.csv")
SalesTrans = SalesTrans %>% inner_join(Location, by = "LocationID")
SalesTrans = SalesTrans %>% inner_join(MerGroup, by = "MerGroup")
LocationID = as.factor(SalesTrans$LocationID)
SalesTrans$ProductID = as.factor(SalesTrans$ProductID)
SalesTrans$Description = as.factor(SalesTrans$Description)
SalesTrans$MerGroup = as.factor(SalesTrans$MerGroup)

# breaking out Q4 to simplify exercise

SalesTrans$Qtr = quarter(SalesTrans$Tdate)
SalesTrans = filter(SalesTrans, Qtr == 4)

```

```{r, echo = F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}
SalesTransSummary = SalesTrans %>% 
  group_by(Description, Population, MerGroup, MfgPromo, Wk ) %>% 
  summarise(Volume = n(), TotSales = sum(Amount) )

```
### Data and EDA

Load the data from last week  *(Sales Transactions)*, and filter for Q4. Summarize by Description, Population, MerGroup, MfgPromo and Wk. Create a visualization of TotSales ~ Population Group *(converted to factor)*, as follows:
```{r, echo = F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

set.seed(222)

SalesTransSummary$PopGrp = as.numeric(as.factor(SalesTransSummary$Population))

# plot using ggextra format
plot_center = ggplot(SalesTransSummary, aes(x=Population,y=TotSales, colour = factor(PopGrp))) + 
  geom_point(width = .05) +
  geom_smooth(method="lm", se = F) +
  theme(panel.background = element_rect(fill = "white")) +  
#  xlim(1, 8) + ylim(0, 10) + 
  ylab("Sales") + xlab("Population")
p1 = ggMarginal(plot_center, type="density", groupColour = FALSE, groupFill = TRUE)
p1
```
Now show sales distributions by Population Group:  

```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

p2 = ggplot(SalesTransSummary, aes(x = TotSales, fill = factor(PopGrp))) + 
  geom_density(bw = 25000, alpha = .2) +
  xlim(-20000, 200000)  +
  theme(panel.background = element_rect(fill = "white")) 
p2

```
*(note: this is not a normal distribution - take a look at the sn package)*

# Deliverables:

1. Posterior distributions for TotSales: Population Groups 5 and 10. 

```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

# get parameters for total population (use that for prior)
PopModel <- sn.mple(y = SalesTransSummary$TotSales, opt.method = "nlminb")$cp
PopParameters <- cp2dp(PopModel, family = "SN")
exi <- PopParameters[1] # location
eomega <- PopParameters[2] # scale
ealpha <- PopParameters[3] # shape
denPrior = dsn(SalesTransSummary$TotSales, exi, eomega, ealpha)

# create densities for pop group 5:

Pop5Model <- sn.mple(y = filter(SalesTransSummary, PopGrp == 5)$TotSales, opt.method = "nlminb")$cp
Pop5Parameters <- cp2dp(Pop5Model, family = "SN")
exi5 <- Pop5Parameters[1] # location
eomega5 <- Pop5Parameters[2] # scale
ealpha5 <- Pop5Parameters[3] # shape
denLike5 = dsn(SalesTransSummary$TotSales, exi5, eomega5, ealpha5)

# create posterior
post5  = denLike5*denPrior
# normalize
post5   <- post5/sum(post5)
sum(post5)



#N = nrow(SalesTransSummary)

Pop10Model <- sn.mple(y = filter(SalesTransSummary, PopGrp == 10)$TotSales, opt.method = "nlminb")$cp
Pop10Parameters <- cp2dp(Pop10Model, family = "SN")
exi10 <- Pop10Parameters[1] # location
eomega10 <- Pop10Parameters[2] # scale
ealpha10 <- Pop10Parameters[3] # shape
denLike10 = dsn(SalesTransSummary$TotSales, exi10, eomega10, ealpha10)

post10  = denLike10*denPrior
# normalize
post10   <- post10/sum(post10)
sum(post10)


# normalize likelihoods
denLike5   <- denLike5/sum(denLike5)
sum(denLike5)
denLike10   <- denLike10/sum(denLike10)
sum(denLike10)


p3 = ggplot(SalesTransSummary, aes(x = TotSales, y = denLike5)) + 
  geom_line() +
  geom_line(aes(x = TotSales, y = post5), color = "blue") +
  geom_line(aes(x = TotSales, y = denLike10), color = "black") +
  geom_line(aes(x = TotSales, y = post10), color = "red") +
  xlim(0, 100000)  +
  theme(panel.background = element_rect(fill = "white")) 
p3


```
2. Determine probablity of a weekly sales exceeding 20,000 in groups 2 and 6


Take a look at densities:
```{r, echo=F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

# create densities for pop group 2:

Pop2Model <- sn.mple(y = filter(SalesTransSummary, PopGrp == 2)$TotSales, opt.method = "nlminb")$cp
Pop2Parameters <- cp2dp(Pop5Model, family = "SN")
exi2 <- Pop2Parameters[1] # location
eomega2 <- Pop2Parameters[2] # scale
ealpha2 <- Pop2Parameters[3] # shape
denLike2 = dsn(SalesTransSummary$TotSales, exi2, eomega2, ealpha2)

# create posterior
post2  = denLike2*denPrior
# normalize
post2   <- post2/sum(post2)
#sum(post2)



#N = nrow(SalesTransSummary)

Pop6Model <- sn.mple(y = filter(SalesTransSummary, PopGrp == 6)$TotSales, opt.method = "nlminb")$cp
Pop6Parameters <- cp2dp(Pop10Model, family = "SN")
exi6 <- Pop6Parameters[1] # location
eomega6 <- Pop6Parameters[2] # scale
ealpha6 <- Pop6Parameters[3] # shape
denLike6 = dsn(SalesTransSummary$TotSales, exi6, eomega6, ealpha6)

post6  = denLike10*denPrior
# normalize
post6   <- post6/sum(post6)
#sum(post6)


# normalize likelihoods
denLike2   <- denLike2/sum(denLike2)
#sum(denLike2)
denLike6   <- denLike6/sum(denLike6)
#sum(denLike6)


p3 = ggplot(SalesTransSummary, aes(x = TotSales, y = denLike2)) + 
  geom_line() +
  geom_line(aes(x = TotSales, y = post2), color = "blue") +
  geom_line(aes(x = TotSales, y = denLike6), color = "black") +
  geom_line(aes(x = TotSales, y = post6), color = "red") +
  xlim(0, 100000)  +
  theme(panel.background = element_rect(fill = "white")) 
p3


```
Create simulations:

```{r,  echo=F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

# generate some simulations

SalesSim2 = data.frame(Sales = rep(SalesTransSummary$TotSales,times = as.integer(post2*10000)))

p3 = ggplot(SalesSim2, aes(x = Sales)) + 
  geom_histogram(fill = "blue", alpha = .2, binwidth = 2000) +
  theme(panel.background = element_rect(fill = "white")) 

SalesSim6 = data.frame(Sales = rep(SalesTransSummary$TotSales,times = as.integer(post6*10000)))

p3 = p3 + 
  geom_histogram(data = SalesSim6, aes(x = Sales), fill = "red", alpha = .2, binwidth = 2000) +
  theme(panel.background = element_rect(fill = "white")) 
p3

```
Get parameters and run probability functions:


```{r,  echo=F, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

Post2Model <- sn.mple(y = SalesSim2$Sales, opt.method = "nlminb")$cp
Post2Parameters <- cp2dp(Post2Model, family = "SN")
postxi2 <- Post2Parameters[1] # location
postomega2 <- Post2Parameters[2] # scale
postalpha2 <- Post2Parameters[3] # shape

Post6Model <- sn.mple(y = SalesSim6$Sales, opt.method = "nlminb")$cp
Post6Parameters <- cp2dp(Post6Model, family = "SN")
postxi6 <- Post6Parameters[1] # location
postomega6 <- Post6Parameters[2] # scale
postalpha6 <- Post6Parameters[3] # shape

p2 = (1 - psn(20000, postxi2, postomega2, postalpha2))
p6 = (1 - psn(20000, postxi6, postomega6, postalpha6))

dfComp = data.frame(Description = c("PopGrp = 2", "PopGrp = 6"),
Prob = c(p2, p6
))

dfComp

# levelset on neg sales

```




